---
title: "Assignment 1 feedback"
subtitle: "Bio300B"
author: "Richard J. Telford (Richard.Telford@uib.no)"
institute: "Institutt for biovitenskap, UiB"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE, echo =  TRUE}

```

```{r, message = FALSE}
#activate the package
library(tidyverse)

#import file into object
biomass <- read_delim(file = "data/biomass2015_H.csv",     
                    delim = ",")

#tidy data
biomass_long <- biomass |>
  pivot_longer(cols = H1:H10, 
               names_to = "height", 
               values_to = "value",
               values_drop_na = TRUE)

#mean height with 10 or more observations
biomass_mean_height <- biomass_long |>                     
  group_by(species) |> 
  filter(n() >= 10) |> 
  summarise(mean_height = mean(value)) |> 
  arrange(desc(mean_height))
```
---
# Result

```{r, message = FALSE, echo = FALSE}
biomass_mean_height
```


---
# Downloading the data

In browser "file" > "save as" or with

```{r, eval = FALSE}
download.file(url =  "https://raw.githubusercontent.com/biostats-r/biostats/main/WorkingInR/data/biomass2015_H.csv", 
              destfile = "data/biomass2015_H.csv")
```

---

# Importing the data

```{r, eval = FALSE}
biomass <- read_delim(file = "data/biomass2015_H.csv", delim = ",") # best

read.delim() # not ideal
# Use Rstudio projects with relative paths
setwd("___") # bad idea
read_csv("my drive/Biostatistics/datalab 2/biomass2015_H.csv") # bad idea
file.choose() # very bad idea - need to find the file every time
```
 
---
# pivot longer

```{r}
# very hard to do without pivot longer
biomass %>%
  drop_na(H1:H10) %>% 
  group_by(species) %>% 
  summarise(mean_value = mean(H1:H10))
```

 
---
# Removing the NA

Must be done in or after pivot_longer()

```{r}
biomass %>%   
  drop_na(H1:H10) %>%      #Remove all NA values
  pivot_longer(cols = H1:H10, names_to = "Height", values_to = "value") |> 
  summarise(n = n())
```
```{r}
biomass %>%   
  pivot_longer(cols = H1:H10, names_to = "Height", values_to = "value") |> 
  drop_na(value) |> 
  summarise(n = n())
```

---
##  values_drop_na
```{r}


mb <- microbenchmark::microbenchmark(
  IN =  biomass |>
  pivot_longer(cols = H1:H10, 
               names_to = "height", 
               values_to = "value",
               values_drop_na = TRUE),
 AFTER = biomass |>
  pivot_longer(cols = H1:H10, 
               names_to = "height", 
               values_to = "value") |> 
  drop_na(value)
)

mb
```

---
```{r, message = FALSE}
autoplot(mb)
```


---
# filter fun


```{r, eval=FALSE}
Biomass_long |> drop_na(value) # easiest - best
Biomass_long |> filter(!is.na(value)) # good


Biomass_long |> filter(value != "NA") # accidentally works
Biomass_long |> filter(value >= 0) # accidentally works

Biomass_long |> na.omit() # dangerous
```


---
# dropping rare taxa

`n()`
or `add_count()` AFTER `group_by()`

```{r}
# with mutate
biomass_long |> 
  group_by(species) |>  
  mutate(n = n()) |>  
  filter(n< 10)

# one step
biomass_long |> 
  group_by(species) |>  
  filter(n() < 10)

# with summarise (very hard)
biomass_long |> 
  group_by(species) |>  
  summarise(n = n()) |>  
  filter(n< 10)
```

---

```{r}
#gjsnitt høyde til de med mer enn 10 
summarise(group_by(biomass_long, species),
          mean(value, na.rm= TRUE)) %>% 
  filter(species != ("Agrostis")) %>% 
  filter(species != ("Fragaria orientalis")) %>% 
  filter(species != ("Gentiana trichotoma ")) %>%  
  filter(species != ("Halenia elliptica")) %>%  
  filter(species != ("Juncus leucomelas")) %>%  
  filter(species != ("Ligusticum scapiforme")) %>%  
  filter(species != ("Pedicularis.sp")) %>% 
  filter(species != ("Primula amethystina subsp.argutidens")) %>% 
  filter(species != ("mix")) %>% 
  filter(species != ("Rhodiola bupleuroides"))
```

---
# select()?

```{r}
mb <- microbenchmark::microbenchmark(
  BEFORE =  biomass |> 
  select(species, H1:H10) |> 
  pivot_longer(cols = H1:H10, 
               names_to = "height", 
               values_to = "value"),
 AFTER = biomass |>
  pivot_longer(cols = H1:H10, 
               names_to = "height", 
               values_to = "value") |> 
  select(species, value)
)

mb

```



---
# comments

- long comments at ends of lines are difficult to read.
- Put long comments on line before code

---
# weeds

Remove unused code

```{r, eval = FALSE}
# All the height column which have 10 readings (removed all NA)
View(height_all)
names(height_all)
str(height_all)
height_all_long <- height_all |> pivot_longer(cols = H1:H10, names_to = "heights", values_to = "height")
view(height_all_long)

```

---
# Missing code

include all code necessary to reproduce result

- library()
- import code


---
# pipes

new line after every pipe.

```{r, eval = FALSE}
# Difficult to read without pipes
summarise(group_by(biomass, species), mean = mean(height, na.rm = TRUE))
```



