---
title: "Multiple Regression"
subtitle: "Bio300B Lecture 8"
author: "Richard J. Telford (Richard.Telford@uib.no)"
institute: "Institutt for biovitenskap, UiB"
date: today
date-format: "D MMMM YYYY"
format: 
  revealjs:
    theme: [default, css/custom.scss]
    chalkboard: true
    resources: 
      - shinylive-sw.js
filters:
  - shinylive
execute: 
  echo: true
  warning: false
  error: true
---

## Outline

```{r}
#| label: setup
#| echo: false
#| message: false
library("tidyverse")
library("patchwork")
library(broom)
library(glue)
theme_set(theme_bw(base_size = 18))
data(penguins, package = "palmerpenguins")
options(dplyr.print_min = 5, dplyr.print_max = 5, digits = 4)
```


- Types of models
- Interactions
- Model selection
  - Exploratory data analysis
  - Anova for hypothesis testing
- Multi-collinearity
- Autocorrelation
- Reporting statistics

## Adding more variables

```{r}
# Two way anova - two categorical predictors
mod_anova2 <- lm(bill_length_mm ~ sex + species, data = penguins)

# ancova - one categorical one continuous predictor
mod_ancova <- lm(bill_length_mm ~ sex + flipper_length_mm, data = penguins)

# Multiple regression - two continuous
mod_mult <- lm(bill_length_mm ~ body_mass_g + flipper_length_mm, data = penguins)
```

---

```{r}
summary(mod_anova2)
```

---

```{r}
summary(mod_ancova)
```

---

```{r}
summary(mod_mult)
```


## Different formula

```{r}
#| label: diff-models-reg
#| fig.height: 6.5
#| echo: false
#| fig.width: 10.0
adelie <- penguins |>
  filter(species == "Adelie") |>
  drop_na(bill_length_mm, flipper_length_mm, sex)


base_plot <- ggplot(adelie, aes(x = flipper_length_mm, y = bill_length_mm)) +
  labs(x = "Flipper length mm", y = "Bill length mm", colour = "Sex")

# intercept only
mod0 <- lm(bill_length_mm ~ 1, data = adelie)

# slope
mod_slope <- lm(bill_length_mm ~ flipper_length_mm, data = adelie)

# categorical predictor
mod_cat <- lm(bill_length_mm ~ sex, data = adelie)

# ancova
mod_ancova <- lm(bill_length_mm ~ sex + flipper_length_mm, data = adelie)

# interaction
mod_interact <- lm(bill_length_mm ~ sex * flipper_length_mm, data = adelie)


preds <- bind_rows(
  "y ~ 1" = adelie |> select(-sex) |> mutate(.fitted = coef(mod0[1])),
  "y ~ x" = augment(mod_slope),
  "y ~ z" = left_join(adelie |> drop_na(sex, flipper_length_mm, bill_length_mm), augment(mod_cat)),
  "y ~ x + z" = augment(mod_ancova),
  "y ~ x * z" = augment(mod_interact),
  .id = "model"
) |>
  left_join(tribble(
    ~model, ~type,
    "y ~ 1", "Intercept only",
    "y ~ x", "Continuous predictor",
    "y ~ z", "Categorical predictor",
    "y ~ x + z", "Continuous & categorical predictors",
    "y ~ x * z", "Interaction"
  )) |>
  mutate(
    model = factor(model, levels = c("y ~ 1", "y ~ x", "y ~ z", "y ~ x + z", "y ~ x * z")),
    type = factor(type, levels = c("Intercept only", "Continuous predictor", "Categorical predictor", "Continuous & categorical predictors", "Interaction")),
    sex2 = if_else(
      type %in% c("Intercept only", "Continuous predictor"),
      true = "Both sexes",
      false = sex
    ),
    sex2 = factor(sex2, levels = c("female", "male", "Both sexes"), labels = c("Female", "Male", "Both sexes"))
  )

base_plot +
  geom_point(data = preds, aes(colour = sex2), alpha = 0.5) +
  geom_line(data = preds, aes(y = .fitted, colour = sex2), linewidth = 1.5) +
  facet_wrap(vars(type, model)) +
  scale_color_brewer(palette = "Set1", na.value = "grey30") +
  theme(legend.position = c(5 / 6, 1 / 4)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 4)))
```

---

```{r}
#| label: diff-models-anova
#| fig.height: 8.0
#| echo: false
#| fig.width: 10.0
library(ggbeeswarm)


base_plot <- ggplot(penguins |> drop_na(sex, bill_length_mm), aes(x = "Penguin", y = flipper_length_mm)) +
  geom_violin() +
  geom_quasirandom(alpha = 0.3, dodge.width = 1) +
  labs(x = "Species", y = "Flipper length mm") +
  scale_color_brewer(palette = "Set1") +
  theme(axis.title.x = element_blank())

# intercept only
mod0 <- lm(flipper_length_mm ~ 1, data = penguins)

# anova
mod_anova <- lm(flipper_length_mm ~ species, data = penguins)

# 2 way anova
mod_anova2 <- lm(flipper_length_mm ~ species + sex, data = penguins)

# 2 way anova with interaction
mod_interaction <- lm(flipper_length_mm ~ species * sex, data = penguins)



p1 <- base_plot +
  geom_pointrange(
    aes(y = .fitted, ymax = .upper, ymin = .lower),
    data = augment(mod0, interval = "confidence") |> slice(1),
    colour = "red",
    size = 1.1
  ) +
  ggtitle("Intercept only", subtitle = "y ~ 1")



p2 <- base_plot +
  aes(x = species) +
  geom_pointrange(
    aes(y = .fitted, ymax = .upper, ymin = .lower),
    data = augment(mod_anova, interval = "confidence", newdata = distinct(penguins, species)),
    colour = "red",
    size = 1.1
  ) +
  ggtitle("Anova", subtitle = "y ~ x")


p3 <- base_plot +
  aes(x = species, colour = sex) +
  geom_pointrange(
    aes(y = .fitted, ymax = .upper, ymin = .lower, colour = sex),
    data = augment(mod_anova2, interval = "confidence", newdata = distinct(penguins, species, sex) |> drop_na(sex)),
    position = position_dodge(width = 1),
    size = 1.1
  ) +
  ggtitle("Two-way anova", subtitle = "y ~ x + z")


p4 <- base_plot +
  aes(x = species, colour = sex) +
  geom_pointrange(
    aes(y = .fitted, ymax = .upper, ymin = .lower, colour = sex),
    data = augment(mod_interaction, interval = "confidence", newdata = distinct(penguins, species, sex) |> drop_na(sex)),
    position = position_dodge(width = 1),
    size = 1.1
  ) +
  ggtitle("Interaction", subtitle = "y ~ x * z OR y ~ x + z + x:z")


p1 + p2 + p3 + p4 & theme(legend.position = "none")
```

## What is an interaction

Effect of one predictor depends on value of another

```{r}
#| label: interaction
#| fig.width: 8.0
#| fig.height: 6.0
#| echo: false

p1 <- crossing(species = c("A", "B"), sex = c("F", "M")) |>
  mutate(
    mean = if_else(species == "A", 0.9, 2.1),
    mean = if_else(sex == "F", mean, mean + 1)
  ) |>
  ggplot(aes(x = species, y = mean, ymin = mean - 0.5, ymax = mean + 0.5, colour = sex)) +
  geom_pointrange(position = position_dodge(width = .5)) +
  labs(y = "Value", title = "No Interaction", x = "Predictor 1", colour = "Predictor 2")

p2 <- crossing(species = c("A", "B"), sex = c("F", "M")) |>
  mutate(
    mean = if_else(species == "A", 0.9, 2.1),
    mean = if_else(sex == "F", mean, mean + 1),
    mean = if_else(species == "B" & sex == "M", mean + 0.5, mean)
  ) |>
  ggplot(aes(x = species, y = mean, ymin = mean - 0.5, ymax = mean + 0.5, colour = sex)) +
  geom_pointrange(position = position_dodge(width = .5)) +
  labs(y = "Value", title = "Interaction", x = "Predictor 1", colour = "Predictor 2")

p1 + p2 + plot_layout(guides = "collect")
```

## Interaction between categorical predictors

:::: {.columns}

::: {.column width="90%"}
```{r}
mod <- lm(flipper_length_mm ~ species * sex, data = penguins)
summary(mod)
```
:::

::: {.column width="20%"}
```{r}
#| label: species_sex_model
#| echo: false
penguins |>
  drop_na(sex) |>
  ggplot(aes(x = species, y = flipper_length_mm, fill = sex)) +
  geom_violin(draw_quantiles = 0.5) +
  labs(x = "Species", y = "Flipper length mm", fill = "Sex")
```

:::

::::

## Understanding coefficients


```{shinylive-r}
#| label: coef-explain-app
#| standalone: true
#| viewerHeight: 650

library(shiny)
library(bslib)

ui <- page_sidebar(
  sidebar = sidebar(checkboxGroupInput("pred",
    "Predictor",
    choiceValues = c(
      "Food", "Temperature",
      "Interaction"
    ), choiceNames = c(
      "Food (A vs. B)",
      "Temperature (Low vs. High)", "Interaction"
    )
  )), card(
    textOutput("formula")
  ), card(
    card_header("Coefficients"),
    tableOutput("coef_table")
  ), card(plotOutput("plot")),
  tags$head(tags$style("#coef_table td{\n                     position:relative;\n                     };\n\n                     ")),
)
server <- function(input, output, session) {
  data <- reactive({
    b0 <- 5
    b1 <- 3
    b2 <- 2
    b3 <- 2
    food <- factor(rep(c("A", "B"), times = 24))
    temperature <- factor(rep(c("Low", "High"), each = 24),
      levels = c("Low", "High")
    )
    response <- b0 + b1 * (food == "B") + b2 * (temperature ==
      "High") + b3 * (food == "B" & temperature ==
      "High") + rnorm(48)
    data.frame(
      food = food, temperature = temperature,
      response = response
    )
  })
  form <- "response ~"
  form2 <- reactive({
    if ("Interaction" %in% input$pred) {
      paste(form, "food * temperature")
    } else if ("Food" %in% input$pred & "Temperature" %in%
      input$pred) {
      paste(form, "food + temperature")
    } else if ("Food" %in% input$pred) {
      paste(form, "food")
    } else if ("Temperature" %in% input$pred) {
      paste(form, "temperature")
    } else {
      paste(form, "1")
    }
  })
  observe({
    if ("Interaction" %in% input$pred) {
      updateCheckboxGroupInput(session, "pred", selected = c(
        "Food",
        "Temperature", "Interaction"
      ))
    }
  })
  model <- reactive({
    lm(form2(), data = data())
  })
  coefs <- reactive({
    coef(model())
  })
  coef_colours <- reactive({
    c(
      `(Intercept)` = "skyblue", foodB = "red", temperatureHigh = "green",
      `foodB:temperatureHigh` = "orange"
    )[names(coefs())]
  })
  coef_table <- reactive({
    c1 <- "<div style=\"width: 100%; height: 100%; z-index: 0; background-color: "
    c2 <- "; position:absolute; top: 0; left: 0; padding:5px;\">\n<span>"
    c3 <- "</span></div>"
    tab <- data.frame(
      Beta = paste0(
        c1, coef_colours(),
        c2, "β", seq_along(coefs()), c3
      ), Coefficent = names(coefs()),
      Estimate = coefs()
    )
    tab
  })
  output$formula <- renderText(paste("Model formula:", form2()))
  output$coef_table <- renderTable(coef_table(), sanitize.text.function = function(x) x)
  output$plot <- renderPlot({
    set.seed(1)
    f <- as.formula(form2())
    fc <- as.character(f)
    ylim <- c(0, max(data()$response))
    par(mar = c(1.5, 2.5, .5, .5), mgp = c(1.5, 0.5, 0))
    if (fc[3] == "1") {
      stripchart(data()$response,
        method = "jitter",
        jitter = 0.1, vertical = TRUE, pch = 1, ylim = ylim,
        ylab = "response"
      )
    } else {
      stripchart(f,
        data = data(), method = "jitter",
        jitter = 0.1, vertical = TRUE, pch = 1, ylim = ylim
      )
    }
    cols <- c("food", "temperature")[c(grepl(
      "food",
      fc[3]
    ), grepl("temperature", fc[3]))]
    if (length(cols) == 0) {
      cols <- "food"
    }
    pred <- predict(model(), newdata = unique(data()[,
      cols,
      drop = FALSE
    ]))
    points(seq_along(pred), pred,
      col = "#832424", pch = 16,
      cex = 4
    )
    xs <- seq_along(pred) - 0.2
    arrows(xs, rep(0, length(pred)), xs, rep(
      coefs()[1],
      length(pred)
    ),
    col = coef_colours()[1], lwd = 4,
    length = 0.1
    )
    pos <- 2 - 0.2
    if ("foodB" %in% names(coefs())) {
      arrows(pos, coefs()[1], pos, coefs()[1] + coefs()["foodB"],
        col = coef_colours()["foodB"], lwd = 4, length = 0.1
      )
      pos <- pos + 1
    }
    if ("temperatureHigh" %in% names(coefs())) {
      arrows(pos, coefs()[1], pos, coefs()[1] + coefs()["temperatureHigh"],
        col = coef_colours()["temperatureHigh"], lwd = 4,
        length = 0.1
      )
      pos <- pos + 1
    }
    if (all(c("foodB", "temperatureHigh") %in% names(coefs()))) {
      arrows(pos, coefs()[1], pos, coefs()[1] + coefs()["foodB"],
        col = coef_colours()["foodB"], lwd = 4, length = 0.1
      )
      arrows(pos, coefs()[1] + coefs()["foodB"], pos,
        coefs()[1] + coefs()["foodB"] + coefs()["temperatureHigh"],
        col = coef_colours()["temperatureHigh"], lwd = 4,
        length = 0.1
      )
      if ("foodB:temperatureHigh" %in% names(coefs())) {
        arrows(pos, coefs()[1] + coefs()["foodB"] +
          coefs()["temperatureHigh"], pos, coefs()[1] +
          coefs()["foodB"] + coefs()["temperatureHigh"] +
          coefs()["temperatureHigh"],
        col = coef_colours()["foodB:temperatureHigh"],
        lwd = 4, length = 0.1
        )
      }
    }
  })
}
shinyApp(ui, server)

```


## Interaction between continuous and categorical predictors

```{r}
#| label: cont-int
#| echo: false
#| fig.width: 10.0
#| fig.height: 6.0

tribble(
  ~type, ~predictor, ~intercept, ~slope,
  "No interaction", "A", 10, 5,
  "No interaction", "B", 12, 5,
  "Interaction - steeper", "A", 10, 5,
  "Interaction - steeper", "B", 12, 8,
  "Interaction - shallower", "A", 10, 5,
  "Interaction - shallower", "B", 12, 3,
) |>
  crossing(continuous = 1:100) |>
  mutate(
    value = continuous * slope / 50 + intercept,
    type = factor(type),
    type = fct_rev(type)
  ) |>
  ggplot(aes(x = continuous, y = value, colour = predictor)) +
  geom_line() +
  facet_wrap(~type) +
  labs(x = "Continuous Predictor", colour = "Categorical\npredictor", y = "Response")
```

---

```{r}
mod_interact <- lm(bill_length_mm ~ sex * flipper_length_mm, data = adelie)
summary(mod_interact)
```



## Interaction between two continuous predictors

```{r}
#| label: cont-cont
#| echo: false
#| fig.width: 8.0
#| fig.height: 6.0
crossing(A = 1:100, B = 1:100, type = c("Interaction", "No interaction")) |>
  mutate(
    type = factor(type),
    type = fct_rev(type),
    response = if_else(type == "Interaction", (A * 7 + B * 5 + A * B * 2) / 10, A * 7 + B * 5)
  ) |>
  ggplot(aes(x = A, y = B, z = response)) +
  geom_contour_filled() +
  facet_wrap(~type) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Predictor A", y = "Predictor B", fill = "Response")
```

---

```{r}
mod <- lm(flipper_length_mm ~ body_mass_g * bill_length_mm, data = adelie)
summary(mod)
```

## Power needed for interactions

- Main effect = $\bar{y_1}$ - $\bar{y_2}$
- Interaction = ($\bar{y_1}$ - $\bar{y_2}$) - ($\bar{y_3}$ - $\bar{y_4}$)

```{r}
#| label: power-int
#| echo: false
p2
```

---

```{r}
set.seed(42)
sigma <- 10
n <- 1000
random <- tibble(
  x = sample(c("A", "B"), n, replace = TRUE),
  z = sample(factor(1:2), n, replace = TRUE),
  y = rnorm(n, mean = 0, sd = sigma)
)

lm(y ~ x * z, data = random) |> tidy()
```

## Formula for interactions

y ~ x + z + x:z

y ~ x * z

y ~ (x + z)^2

Use y ~ x + I(x^2) to get a quadratic. Or y ~ poly(x, 2)

---

```{r}
mod1 <- lm(flipper_length_mm ~ species + sex + species:sex, data = penguins)
mod2 <- lm(flipper_length_mm ~ species * sex, data = penguins)
mod3 <- lm(flipper_length_mm ~ (species + sex)^2, data = penguins)

coef(mod1)
coef(mod2)
coef(mod3)
```

## Model Selection

You want to the best model!

The best model for what?

- Exploratory data analysis
- Inference (hypothesis testing)
- Predictions

::: aside
Tredennick et al 2021 A practical guide to selecting models for exploration, inference, and prediction in ecology Ecology. [https://doi.org/10.1002/ecy.3336]( https://doi.org/10.1002/ecy.3336) 
:::

## Exploratory analysis

- Consider all plausible models
- P-values **not** meaningful
- High type I error rate
- Suggest hypotheses for hypothesis testing **with independent data**

The more biology you include in the model 

## Automatic model selection

Last resort - many problems and biases

- Forward selection

- Backwards selection

- All possible models

---

```{r}
#| label: set-options
#| echo: false
old <- options(width = 120)
```


```{r}
library(MuMIn)
library(conflicted)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

penguins2 <- penguins |>
  drop_na() |>
  select(-year)
full <- lm(body_mass_g ~ ., data = penguins2, na.action = "na.fail")
```

## AIC

AIC Akaike information criterion $2k - 2 \times log(likelihood)$  
	(_k_ the number of parameters)
	
Measure of how well the model fit the data


Penalised for model complexity


AICc correction for small sample sizes

AIC weights - probability model is best of those tested

---

```{r}
dredge(full) |> gt::gt()
```


```{r}
#| label: reset-options
#| echo: false
options(old)
```

## Inference

Test small number of _a priori_ hypothesis.

Use `anova()` to compare nested models

H<sub>0</sub> there is no interaction between sex and species for predicting body mass

```{r}
mod1 <- lm(body_mass_g ~ species + sex, data = penguins)
mod2 <- lm(body_mass_g ~ species * sex, data = penguins)
anova(mod1, mod2)
```

## car::Anova vs anova


```{r}
mod2 <- lm(body_mass_g ~ species + sex, data = penguins)
mod3 <- lm(body_mass_g ~ sex + species, data = penguins)
```

- anova - tests terms sequentially
- car::Anova - marginal test

---

```{r}
anova(mod2)
anova(mod3)
```

---

```{r}
car::Anova(mod2)
car::Anova(mod3)
```


## Multicollinearity

- Two or more predictor variables in a 
multiple regression model are highly 
correlated. Example: pH and 
calcium
- Coefficient estimates are unstable 
  - erratic change in response to small 
changes in the model or the data.

- Solve by having lots of data

--- 

```{r}
#| label: collinearity
col <- performance::check_collinearity(full)
col
plot(col)
```

## Autocorrelation

Linear models assume residuals are independent

If data are spatially or temporally structured, residuals may be correlated.

Positive autocorrelation 

- confidence intervals are too narrow
- increases risk of false positive (Type I error) 


## Luteinising Hormone concentration

```{r}
#| label: autocorrelation
#| echo: true
#| code-fold: true

library(ggfortify)

lh <- fortify(lh) |>
  mutate(time = Index * 10) |>
  rename(concentration = Data)

mod_lh <- lm(concentration ~ time, data = lh)

augment(mod_lh, interval = "confidence") |>
  ggplot(aes(x = time, y = concentration)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.2, fill = "#ee5050") +
  geom_line(aes(y = .fitted), colour = "#ee5050") +
  labs(x = "Time, minutes", y = "Hormone concentration")
```

## Detecting autocorrelation

For time series with equally-spaced observations, use autocorrelation function (ACF)

```{r}
autoplot(acf(mod_lh$residuals, plot = FALSE))
```

## Detecting autocorrelation

Tests for equally-spaced observations

Durbin-Watson test

```{r}
performance::check_autocorrelation(mod_lh)
```

Other methods for spatial data and non-equally spaced data

Solution - use a model that accounts for autocorrelation

- generalised least squares (`nlme::gls()`)


## Reporting regression results - tables

Make a table - estimates, confidence intervals, p-values

Can calculate confidence intervals with `broom::tidy()` and format output

```{r}
library(gt)
tidy(mod_mult, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  mutate(p.value = format.pval(p.value, eps = 0.001)) |>
  gt() |>
  fmt_number(
    columns = c(estimate, conf.low, conf.high),
    decimals = 4
  )
```

## `gtsummary`

Or make table with `gtsummary`

```{r}
library(gtsummary)
tbl <- tbl_regression(mod_mult,
  label = list(
    body_mass_g = "Body mass g",
    flipper_length_mm = "Flipper length mm"
  ),
  estimate_fun = label_style_sigfig(digits = 4)
)
tbl
```

## Reporting regression results - inline

Can extract values directly from regression . 

- estimate
- confidence intervals
- p-value, degrees of freedom, statistic

```{markdown}
Longer flipper length is associated with longer bill length (estimate = `r round(coef(mod_mult)["flipper_length_mm"], 2)` mm/mm, 95% CI = `r round(confint(mod_mult)["flipper_length_mm", ], 2)` p = `r format.pval(tidy(mod_mult) |> filter(term == "flipper_length_mm") |> select("p.value"), eps = 0.001)`).
```

Longer flipper length is associated with longer bill length (estimate = `r round(coef(mod_mult)["flipper_length_mm"], 2)` mm/mm, 95% CI = `r round(confint(mod_mult)["flipper_length_mm", ], 2)` p = `r format.pval(tidy(mod_mult) |> filter(term == "flipper_length_mm") |> select("p.value"), eps = 0.001)`).


Also function `inline_text()` from `gtsummary`

See APA guide for reporting statistics
