---
title: "Mixed effect Models"
subtitle: "Bio300B Lecture 10"
author: "Richard J. Telford (Richard.Telford@uib.no)"
institute: "Institutt for biovitenskap, UiB"
date: today
date-format: "D MMMM YYYY"
format: 
  revealjs:
    theme: [default, css/custom.scss]
    chalkboard: true
execute: 
  echo: true
  warning: false
  error: true
---


## Assumptions of Least Squares 

```{r setup, echo=FALSE, message=FALSE}
library("tidyverse")
library("patchwork")
library(broom)
library(glue)
library(ggbeeswarm)
library(faux)
library(lme4)
library(patchwork)
theme_set(theme_bw(base_size = 18))
data(penguins, package = "palmerpenguins")
#options(dplyr.print_min = 2, dplyr.print_max = 3, digits = 4)
```

1. Linear relationship between response and predictors.
2. The residuals have a mean of zero.
3. The residuals have constant variance (not heteroscedastic).
4. [**The residuals are independent (uncorrelated).**]{style="color: red;"}
5. The residuals are normally distributed.

- Spatial/temporal/phylogenetic autocorrelation
- Clustered/hierarchical data


## Clustered data{background-image="figures/46695412244_66616e8523_c.jpg"}

Observations not independent

- clustered data
- repeat measurements

Ignoring it causes pseudoreplication

::: aside
[WorldFish](https://www.flickr.com/photos/theworldfishcenter/46695412244)
:::

## Two tanks - bad design

```{r two-tanks-design, echo = FALSE}
#simulate with faux package

# define parameters
subj_n = 10  # number of subjects
item_n = 10  # number of items
b0 = 1000    # intercept
b1 = 100       # fixed effect of condition
u0s_sd = 100   # random intercept SD for subjects
sigma_sd = 100 # error SD


set.seed(42)
demo <- add_random(subj = subj_n, tank = 2) |> 
  add_between("tank",  condition = c("control", "test")) |> 
  add_recode("condition", "cond.t", control = 0, test = 1) |> 
  add_ranef("tank", u0 = u0s_sd) |> 
  add_ranef(sigma = sigma_sd) |> 
  # calculate DV
  mutate(Mass_g = b0 +  u0 + (b1) * cond.t + sigma)

demo |> 
  ggplot(aes(x = tank, y = Mass_g)) +
  geom_violin() +
  geom_quasirandom(aes(colour = condition))
```


## Two tanks - bad models

```{r}
mod_bad <- lm(Mass_g ~ condition, data = demo)
mod_bad2 <- lm(Mass_g ~ tank, data = demo)

mod_bad
mod_bad2
```

## Better design - replicate tanks

```{r many-tanks-design, echo = FALSE}
set.seed(42)
demo2 <- add_random(subj = subj_n, tank = 8) |> 
  add_between("tank",  condition = c("control", "test")) |> 
  add_recode("condition", "cond.t", control = 0, test = 1) |> 
  add_ranef("tank", u0 = u0s_sd) |> 
  add_ranef(sigma = sigma_sd) |> 
  # calculate DV
  mutate(Mass_g = b0 +  u0 + (b1) * cond.t + sigma)

outer_plot <- demo2 |> 
  ggplot(aes(x = tank, y = Mass_g)) +
  geom_violin() +
  geom_quasirandom(aes(colour = condition))

outer_plot
```


## Better design, bit better model

```{r}
mod_bit <- lm(Mass_g ~ condition + tank, data = demo2)
summary(mod_bit)

```


## Fixed effect and random effects

### <font color = "red">Fixed effect</font> effect model

$$y_i = \color{red}{\beta_0} + \color{red}{\beta_1}x_i + \color{red}{\beta_2}tank_i  +\color{blue}{ \epsilon_i}$$

Not interested in effect of tank 

### Mixed effect model

Use a <font color = "blue">Random effect</font> 

Assumes observed tanks from a population of possible tanks
 
$$y_{ij} = \color{red}{\beta_0} +\color{blue}{b_{0i}}+ \color{red}{\beta_1}x_i +\color{blue}{ \epsilon_{ij}}$$

## Random effects

Residuals from a normal distribution $\color{blue}{ \epsilon_{ij}} \sim N(0, \sigma_{ind})$  
Random effects from a normal distribution $\color{blue}{b_{0i}} \sim N(0, \sigma_{clu})$

```{r control, echo = FALSE, fig.height = 5}
p1 <- demo2 |> 
  filter(condition == "control") |> 
  ggplot(aes(x = tank, y = Mass_g)) +
  geom_violin() +
  geom_quasirandom(aes(colour = condition)) +
  stat_summary(fun = "mean", geom= "point", size = 2) +
  theme(legend.position = "none")

p2 <- tibble(x = -350:350, y = dnorm(x, mean = 0, sd = u0s_sd)) |> 
  ggplot(aes(x, y)) +
  geom_area() +
  coord_flip() +
  labs(x = "Mass g", y = "Density")

p1 + p2
```

## Fitting a mixed effect model

```{r}
library(lme4)
mod1 <- lmer(Mass_g ~ condition + (1|tank), data = demo2)
```

---

```{r}
summary(mod1)
```


---

```{r, message = FALSE}
library(lmerTest)
mod1a <- lmer(Mass_g ~ condition + (1|tank), data = demo2)
summary(mod1a)
```

## Alternative package

```{r, message = FALSE}
library(nlme)
mod2 <- lme(Mass_g ~ condition, data = demo2, random =  ~ 1|tank)
summary(mod2)
```


## Fixed or random effects{.scrollable}

Fixed effects factors:

- Informative factor levels with regard to the hypothesis
- Effect of each level interesting
- The levels are deliberately chosen by experimenter
- Increasing sample size doesn't increase number of levels 

Random effects factors:

- Uninformative factor levels with regard to the hypothesis
- Effect of each level uninteresting
- Increasing sample size often increases number of levels
- Part of a population of possible levels

---

"one modeler’s random effect is another modeler’s fixed effect."

"Are there enough levels of the factor in the data on which to base an estimate of the variance of the population of effects? No, means fixed effects."

::: aside
Bolker B (2021) [GLMM FAQ](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html)
:::


## Fixed or random effects?

- snake id number
- vaccinated or not
- species
- light level
- litter of puppies
- site


## Outer and inner variables

:::: {.columns}

::: {.column width="50%"}

```{r inner-outer, echo = FALSE, fig.height = 6.5, fig.width=5}
set.seed(42)
demo3 <- add_random(subj = subj_n/2, tank = 8) |> 
  add_within("tank",  condition = c("control", "test")) |> 
  add_recode("condition", "cond.t", control = 0, test = 1) |> 
  add_ranef("tank", u0 = u0s_sd) |> 
  add_ranef(sigma = sigma_sd) |> 
  # calculate DV
  mutate(Mass_g = b0 +  u0 + (b1) * cond.t + sigma)

inner_plot <- outer_plot %+% demo3 

(outer_plot + ggtitle("Outer variable")) /
(inner_plot + ggtitle("Inner variable")) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```
:::

::: {.column width="50%"}
- which design is more powerful?
- which design is practical
:::

::::

---

```{r}
mod3 <- lmer(Mass_g ~ condition + (1|tank), data = demo3)
summary(mod3)
```

## Continuous inner predictor

```{r sleepstudy, fig.height = 5, echo = 1}
data("sleepstudy", package = "lme4")
sleepstudy |> filter(Days >= 2) |> 
ggplot(aes(x = Days, y = Reaction, colour = Subject)) +
  geom_point() +
  labs(x = "Days of sleep deprivation", y = "Average reaction time (ms)")
```

## Random intercept


```{r random-intercept}
fm1 <- lmer(Reaction ~ Days + (1|Subject), sleepstudy, subset = Days>=2)

summary(fm1)

```

---

```{r random-intercept-plot, fig.height=5}
library(broom.mixed)
augment(fm1) |> 
  ggplot(aes(x = Days, y = Reaction, colour = Subject)) +
  geom_point() +
  geom_line(aes(y = .fitted)) +
  geom_line(aes(y = .fixed), colour = "black", size = 1.5) +
  theme(legend.position = "none")
  
```



## Random slope


```{r random-slope}
fm2 <- lmer(Reaction ~ Days + (Days|Subject), sleepstudy, subset = Days>=2)

summary(fm2)

```

---

```{r random-slope-plot, echo = FALSE}
augment(fm2) |> 
  ggplot(aes(x = Days, y = Reaction, colour = Subject)) +
  geom_point() +
  geom_line(aes(y = .fitted)) +
  geom_line(aes(y = .fixed), colour = "black", size = 1.5) +
  theme(legend.position = "none")
  
```


## Comparing models

```{r}
anova(fm1, fm2)
```

## Model diagnostics

See ?plot.merMod

```{r diag1}
fm1F <- augment(fm1)
## standardized residuals versus fitted values
ggplot(fm1F, aes(x = .fitted, y = .resid)) + 
  geom_point(colour = "blue") +
  geom_hline(yintercept = 0)
```

---

```{r diag2}
## box-plots of residuals by Subject
ggplot(fm1F, aes(x = Subject, y = .resid)) + 
  geom_boxplot() + 
  coord_flip()
```

---

```{r diag3}
ggplot(fm1F, aes(x = .fitted, y = Reaction)) + 
  geom_point(colour = "blue") +
    facet_wrap(facets = vars(Subject)) +
  geom_abline(intercept = 0, slope = 1)
```

---

```{r diag4}
#qq plot of residuals
library(lattice)
qqmath(fm1)
```

---

```{r diag5}
#qq plot of random effects
qqmath(ranef(fm1))
```


## Crossed random effects

Two random effects

Eggs from birds nests (first random effect - lay_nest) moved to
other nests (second random effect - hatch_nest)

```{r, eval = FALSE}
lmer(mass ~ treatment + (1|lay_nest) + (1|hatch_nest), data = data)
```

## Nested random effects

Hierarchical random effects

- Plot nested in site
- sample nested in fish nested in tank
- species nested in family

```{r, eval = FALSE}
lmer(mass ~ treatment + (1|site/plot), data = data)
```

## Beyond linear mixed effect models

Generalised linear mixed effect models

 - poisson
 - binomial
 
Fit with `glmer()`

Autocorrelated data 

- repeat measurements
- fit with `nlme::lme()`

## Bayesian hierarchial models

Mixed effect models can be hard to fit

- convergence problems

Bayesian model can help

Different statistical philosophy

- rstan
- nimble

Use prior information (or uninformative priors)


## Resources

Bolker B (2021) [GLMM FAQ](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html)

Harrison et al (2018) [A brief introduction to mixed effects modelling and multi-model inference in ecology](https://peerj.com/articles/4794/)


