---
title: "Assignment 6: Mixed effect models"
format: html
editor: visual
---

```{r}
#| label: load-packages
#| output: false
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggbeeswarm)

# set default theme
theme_set(theme_bw())
```

```{r}
#| label: load-process-data
#| output: false
traits <- read_delim(here("data/PFTC4_Svalbard_2018_ITEX_Traits.csv"))

# filter out height data
trait_height <- traits |> 
  filter(Trait == "Plant_Height_cm", Taxon == "bistorta vivipara")
```

```{r}
#| label: fig-schematic-map
#| fig-cap: Simulated maps of _Bistorta vivipara_ plants in the plots. Some plots have no _B. viviara_. 


treatments <-  traits |>
  drop_na(Site) |> 
  distinct(Treatment, Site, PlotID) |> 
  mutate(plot_number = str_extract(PlotID, "\\d{1,2}"),
         plot_number = as.numeric(plot_number)) 

trait_height |> 
  mutate(plot_number = str_extract(PlotID, "\\d{1,2}"),
         plot_number = as.numeric(plot_number),
         x = 0.5, 
         y = 0.5) |> 
  ggplot(aes(x = x, y = y, size = Value)) +
    geom_rect(data = treatments,
              mapping = aes(xmin = -Inf, ymin = -Inf, xmax = Inf, ymax = Inf, fill = Treatment), 
              inherit.aes = FALSE, alpha = 0.4) +
  geom_jitter(alpha = 0.6) + # simulate some x/y positions
  facet_grid(rows =vars(Site), cols =  vars(plot_number)) +
  scale_fill_manual(values = c(OTC = "pink", CTL = "lightblue")) +
  labs(x = "Simulated x position", y = "Simulated y position", size = "Height cm" ) +
  coord_equal() +
  theme(legend.position = "bottom", 
        axis.text = element_blank())
```

```{r}
#| label: run-analyses

# treatment only
mod1 <- lmer(Value ~ Treatment  + (1|PlotID) , data = trait_height)

# treatment and site
mod2 <- lmer(Value ~ Treatment + Site + (1|PlotID) , data = trait_height)

# treatment and site with interaction
mod3 <- lmer(Value ~ Treatment * Site + (1|PlotID) , data = trait_height)

# test if interaction significant
aov_interation <- anova(mod2, mod3) # no evidence of interaction

# test if site significant
aov_site <- anova(mod1, mod2) # no evidence of effect of site/habitat

aov_treatment <- anova(mod1) # weak evidence of effect

summary(mod1)
```

```{r}
#| label: diagnostic-plots


mod1_aug <- augment(mod1)

# standardized residuals versus fitted values
ggplot(mod1_aug, aes(x = .fitted, y = .resid)) + 
  geom_point(colour = "blue") +
  geom_hline(yintercept = 0) 
# looks OK

ggplot(mod1_aug, aes(x = PlotID, y = .resid)) + 
  geom_boxplot() + 
  coord_flip()
# fairly varied, but low number of plants per plot so probably expected

library(lattice)
#qq plot of residuals
qqmath(mod1) #perfect

#qq plot of random effects
qqmath(ranef(mod1)) # one somewhat extreme plot
```

```{r}
#| label: fig-results
#| fig-cap: Effect of warming treatment on height of _Bistorta vivipara_.
augment(mod1) |> 
  ggplot(aes(x = Treatment, y = Value)) +
  geom_violin(colour = "grey40") +
  geom_quasirandom(colour = "grey40") +
  geom_point(aes(y = .fixed), colour = "black", size = 4) + #make large so obvious
  labs(x = "Treatment", y = "Height cm") +
  scale_x_discrete(labels = c(CTL = "Control", OTC = "Open top\nchamber"))

#Could colour/facet by habitat, but not essential
```

Adding confidence intervals is more complex, but can be done with `bootMer` or the `merTools` package.

## Interpretation

```{r}
#| label: extract-results
ci <- confint(mod1)
ci_treatment <- ci["TreatmentOTC", ] |> signif(digits = 2)
```

There is no evidence for an effect of habitat on the height of *B. vivipara* (p = `r  format.pval(aov_site$P[2], digits = 2)`).

There is weak evidence for an effect of the warming treatment on the plant height, with taller plants in the OTCs (mean difference `r round(fixef(mod1)["TreatmentOTC"], 2)` cm, 95 % confidence interval `r paste(ci_treatment, collapse = " -- ")` cm, p = `r format.pval(aov_treatment$P, digits = 2)`).
