---
title: "Datalab 7 - linear regression - solution"
date: today
format: html
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(ggfortify)
data(penguins, package = "palmerpenguins")
theme_set(theme_bw())
```

Please work in an quarto document

## Correlations

- With the Penguins dataset, find the Pearson correlation bill length and bill depth for Gentoo penguins.

```{r}
gentoo <- penguins |> 
  filter(species == "Gentoo")

cor(gentoo$bill_length_mm, gentoo$bill_depth_mm, use = "complete.obs")

#OR
penguins |> 
  filter(species == "Gentoo") |> 
  drop_na(bill_length_mm, bill_depth_mm) |> 
  summarise(corr = cor(bill_length_mm, bill_depth_mm))
```


- Find the confidence intervals of this correlation. Is it significant?

```{r}
ct <- cor.test(gentoo$bill_length_mm, gentoo$bill_depth_mm, use = "complete.obs")
ct
# extracting components with broom
glance(ct) |> select(starts_with("conf"))

# or without
ct$p.value < 0.05
```


- Calculate r^2^. How much of the variation in the data is explained by the correlation.

```{r}
corr <- cor(gentoo$bill_length_mm, gentoo$bill_depth_mm, use = "complete.obs")
corr^2
corr^2 * 100 # %

```


- Find the correlation between bill length and bill depth for
- each species separately 
    
```{r}
penguins |> 
  group_by(species) |> 
  drop_na(bill_length_mm, bill_depth_mm) |> 
  summarise(corr = cor(bill_length_mm, bill_depth_mm))
# all positive
```

    - all species combined
    
```{r}
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use = "complete.obs")
# negative
```

- Make a figure to explain the apparent conflict between these results

```{r}
#| message: false
#| warning: false

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm) +
  geom_smooth(aes(group = 1, colour = "All species"), method = "lm") +
  scale_color_brewer(palette = "Set1", limits = c("All species", "Adelie", "Chinstrap", "Gentoo")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
  
```

This is [Simpson's paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox).

## Continuous predictors

We want to test how bill length varies with body mass in Chinstrap penguins

- make a suitable plot to show the raw data

```{r}
#| label: bill-length-mass-plot
chinstrap <- penguins |> 
  filter(species == "Chinstrap")

ggplot(chinstrap, aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  labs(x = "Body mass g", y = "Bill length mm")
```


- fit a suitable model to the data

```{r}
mod <- lm(bill_length_mm ~ body_mass_g, data = chinstrap)
mod
```

- interpret the model coefficients

For every extra gram body mass, bill length increases by `r coef(mod)[2]` mm.

- is the model statistically significant

```{r}
av <- car::Anova(mod)
av
av$`Pr(>F)`[1] < 0.05
```

- Examine the diagnostic plots. Is everything OK?

```{r}
#| label: diagnostics
#| fig.height: 8.0
autoplot(mod, which = 1:6) # looks good
```


- Predict, with uncertainty, the mean bill length for Chinstraps with a mass of 3000 and 5000 g

```{r}
nd <- tibble(body_mass_g = c(3000, 5000))
augment(mod, newdata = nd, se_fit = TRUE, interval = "confidence")

```

- Add the regression line, with the confidence interval, to the plot.

```{r}
augment(mod, interval = "confidence") |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Body mass g", y = "Bill length mm")
```



## Categorical predictors

We want to test whether Adelie penguins differ in body mass between islands.

- Make a suitable plot to show the raw data

```{r}
#| label: adelie-raw-plot
library(ggbeeswarm)
adelie <- penguins |> filter(species == "Adelie")


island_mass_plot <- ggplot(adelie, aes(x = island, y = body_mass_g)) +
  geom_violin() +
  geom_quasirandom() + 
  labs(x = "Island", y = "Body mass g")
island_mass_plot
```


- Fit a suitable model to the data

```{r}
mod2 <- lm(body_mass_g ~ island, data = adelie)
```


- Interpret the model coefficients

```{r}
coef(mod2) 
coef(mod2)[1] # mean mass on Bisco Island
coef(mod2)[2] # difference in mass between Bisco and Dream
coef(mod2)[3] # difference in mass between Bisco and Torgersen 

```


- Is the model statistically significant?

```{r}
av2 <- car::Anova(mod2)
av2
av2$`Pr(>F)`[1] < 0.05
```


- Include the predicted mass with confidence interval on the plot

```{r}
nd <- distinct(adelie, island)
pred_mass <- augment(mod2, newdata = nd, interval = "confidence")

island_mass_plot +
  geom_pointrange(aes(y = .fitted, ymin = .lower, ymax = .upper), data = pred_mass, colour = "red", size = 1.5)
  
```


- Do a post-hoc test to test for differences between pairs of islands (in real analysis only do this if overall model is significant) (Hint - use `multcomp` package but DO NOT use `library(multcomp)` - it also loads `plyr` and causes conflicts)
- plot the result of the post-hoc test and interpret the results

```{r}
#| label: multcomp-plot

mc <- multcomp::glht(model = mod2, linfct = multcomp::mcp(island = "Tukey"))
mc

plot(mc)
```



## Exploring diagnostic plots

Use the `influence_leverage_app()` from biostats.tutorials package to explore how the diagnostic plots look with a outlier, influential observation, or observation with high leverage.

Open the app and move the red point. Diagnostic plot on the left shows model without the red point, plot on the right includes the red point.

Try to:

- Make the R2 as high as possible
- Make the R2 as low as possible

And look at the different diagnostic plots


