---
title: 'Datalab 8: Generalised linear models'
author: "Richard Telford"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(broom)

```

# Part 1

Relationship between Southern Ocean diatom richness and sea surface temperature.

Data, Esper and Geronde (2014; https://doi.org/10.1016/j.palaeo.2014.08.008), give diatom assemblage composition from surface samples across the Southern Ocean. 
Column diatoms [#] shows the total number of diatoms observed.
Temperature data are not given (could be extracted from an atlas with latitude/longitude), so we will use latitude as a proxy for temperature.

## Download and import data

- Download data as tab-delimited test from https://doi.pangaea.de/10.1594/PANGAEA.828674
- Import the data into R

```{r import-data}
# here() find the file relative to the project root
esper <- read_delim(file = here("data/Esper-Gersonde_2014.tab"), skip = 155, delim = "\t") |> 
  janitor::clean_names()#tidy column names
```



## Process data
- Find the species richness (i.e. number of columns with diatom percent with abundance > 0) (hint: either use pivot_longer() and group_by() and summarise(), or select() and rowSums())

```{r richness}
esper_richness <- esper |> 
  pivot_longer(cols = matches("percent"), names_to = "taxa", values_to = "percent") |> # could also cols = a_actinochilus_percent:diatoms_indet_percent (should probably remove diatoms_indet_percent is it not a real species)
  filter(percent > 0) |> # this will fail if any sites have no diatoms. In that case use summarise(n = sum(percent > 0))
  group_by(event, latitude) |> 
  summarise(n = n())
```



## plot and model data

- plot richness against latitude

```{r richness-plot}
ggplot(esper_richness, aes(x = latitude, y = n)) +
  geom_point() + 
  labs(x = "Latitude °" , y = "Species richness")
```


- fit an appropriate model to find the relationship between richness and latitude

```{r richness-mod}
# count data - try a poisson
mod_richness <- glm(n ~ latitude, data = esper_richness, family = poisson)
mod_richness
```

- plot the fitted values from the model with the raw data. Is the model a good fit. Do you need to add extra terms?

```{r plot-with-fitted}
augment(mod_richness, type.predict = "response") |> 
  ggplot(aes(x = latitude, y = n)) +
  geom_point() +
  geom_line(aes(y = .fitted))

# to get confidence interval
augment(mod_richness, se_fit = TRUE, type.predict = "link") |> 
  mutate(fitted = exp(.fitted), 
         upper = exp(.fitted + 1.96 * .se.fit), 
         lower = exp(.fitted - 1.96 * .se.fit)) |> 
  ggplot(aes(x = latitude, y = n)) +
  geom_point() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3) +
  geom_line(aes(y = fitted))
```
Model appear to fit the data well. Does not appear to need a quadratic over this section of latitude (but probably decreases at high northern latitudes)

Residual deviance < residual df - no evidence of over-dispersion

- interpret the model coefficients

```{r coef-richness}
coef(mod_richness)[1] # log richness when latitude = 0
coef(mod_richness)[2] # positive slope in link space - proportional change in response space
```
```{r}
cr <- coef(mod_richness)
#link space - change of 1 degree
(cr[1] + cr[2] * -60) - (cr[1] + cr[2] * -61)

#reponse space - change of 1 degree
exp(cr[1] + cr[2] * -60) / exp(cr[1] + cr[2] * -61) -1
```


# Part 2

Find the relationship of the presence-absence of _F. cylindrus_  with latitude in the data from above.

```{r f-cylindrus-data}
f_cylindrus_df <- esper |> 
  mutate(f_cylindrus = sign(f_cylindrus_percent))
```


- plot the data

```{r f-cylindrus-plot}
set.seed(1)
p_fc <- ggplot(f_cylindrus_df, aes(x = latitude, y = f_cylindrus)) +
  geom_jitter(width = 0, height = 0.1)

p_fc
```


- fit an appropriate model including a quadratic term (hint: poly())

```{r fc-mod}
mod_fc <- glm(f_cylindrus ~ latitude, data = f_cylindrus_df, family = binomial)
mod_fc  
```

- add the fitted model, with uncertainties, to the plot

```{r fc-plot2}
set.seed(1)
augment(mod_fc, se_fit = TRUE) |> 
  mutate(fitted = plogis(.fitted), 
         upper = plogis(.fitted + 1.96 * .se.fit), 
         lower = plogis(.fitted - 1.96 * .se.fit)) |> 
  ggplot(aes(x = latitude, y = f_cylindrus)) +
  geom_jitter(width = 0, height = 0.1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3) +
  geom_line(aes(y = fitted)) 
```



# Part 3

Find the relationship between the relative abundance of  _F. cylindrus_  with latitude

- plot the data

```{r ra-plot}
# need proportion data rather than percent

f_cylindrus_df <- f_cylindrus_df |> 
  mutate(f_cylindrus_prop = f_cylindrus_percent/100 )

ggplot(f_cylindrus_df, aes(x = latitude, y = f_cylindrus_prop)) +
  geom_point()

```

- fit an appropriate model including a quadratic term (hint: every diatom either is, or is not _F. cylindrus_)

```{r}
# can fit with proportion and weights
mod_ra <- glm(f_cylindrus_prop ~ latitude, 
              data = f_cylindrus_df,
              family = binomial,
              weights = diatoms_number)

# warning message because of rounding errors 
f_cylindrus_df |> 
  filter(f_cylindrus > 0) |> 
  mutate(fc = f_cylindrus_prop * diatoms_number) |> 
  select(f_cylindrus_prop, fc, diatoms_number) # non-integers

# can also fit model with 2 column response: successes and failures
f_cylindrus_df <- f_cylindrus_df |> 
   mutate(fc = f_cylindrus_prop * diatoms_number, 
          fc = round(fc), # round to integer
          not_fc = diatoms_number - fc)

mod_ra2 <- glm(cbind(fc, not_fc) ~ latitude, 
              data = f_cylindrus_df,
              family = binomial)

#result is same (small differences because of fixing the rounding errors)
tidy(mod_ra)
tidy(mod_ra2)

# second method does not play well with augment()

```

- is there evidence of overdispersion?

```{r}
anova(mod_ra) # residual deviance >> residual df
```

- adjust the model if necessary

```{r}
mod_ra_quasi <- glm(f_cylindrus_prop ~ latitude, 
              data = f_cylindrus_df,
              family = quasibinomial,# change to quasibinomial to allow for overdispersion
              weights = diatoms_number)


tidy(mod_ra_quasi) # coefficients are the same
```


- add the fitted model, with uncertainties, to the plot

```{r}
# binomial model
augment(mod_ra, se_fit = TRUE) |> 
  mutate(fitted = plogis(.fitted), 
         upper = plogis(.fitted + 1.96 * .se.fit), 
         lower = plogis(.fitted - 1.96 * .se.fit)) |> 
  ggplot(aes(x = latitude, y = f_cylindrus_prop)) +
  geom_point() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3) +
  geom_line(aes(y = fitted)) 

# with quasibinomial
# line on above plot not smooth because bending too much in region with few data points for predictions. Solve by making predictions for new data on regular grid

nd <- tibble(latitude = seq(min(esper$latitude), max(esper$latitude), length.out = 200))

preds <- augment(mod_ra_quasi, se_fit = TRUE, newdata = nd) |> 
  mutate(fitted = plogis(.fitted), 
         upper = plogis(.fitted + 1.96 * .se.fit), 
         lower = plogis(.fitted - 1.96 * .se.fit)) 

ggplot(f_cylindrus_df, aes(x = latitude, y = f_cylindrus_prop)) +
  geom_point() +
  geom_ribbon(aes(ymin = lower, ymax = upper, y  = NULL), data = preds, alpha = 0.3) +
  geom_line(aes(y = fitted), data = preds) 

# wider ( & more realistic) confidence interval
# much smoother curve
```

