---
title: 'Datalab 8: Multiple regression'
author: "Richard Telford"
date: today
format: html
---

```{r}
#| label: setup
#| message: false

#load packages
library(tidyverse)
library(broom) 
library(ggfortify)
library(conflicted)
conflict_prefer_all("dplyr", quiet = TRUE)

# theme defaults
theme_set(theme_bw())

# load data
data(penguins, package = "palmerpenguins")
```

In an quarto document

# Task 1 2-way anova

- With the penguins data, fit models to predict body mass from species and sex with and without an interaction.

```{r}
mod1 <- lm(body_mass_g ~ sex + species, data = penguins)
mod2 <- lm(body_mass_g ~ sex * species, data = penguins) # * to make the interaction
# OR
#mod2 <- lm(body_mass_g ~ sex + species + sex:species, data = penguins) # explicitly write out interaction - more typing
```


- Interpret the model coefficients.

```{r}
tidy(mod1) # or use summary(mod1) or just coef
# mean Adelie female body mass
# difference between male and female body mass
# difference between chinstrap and adelie
# difference between gentoo and adelie

tidy(mod2)
# mean Adelie female body mass
# difference between male and female adelie body mas
# difference between chinstrap and adelie females
# difference between gentoo and adelie females
# excess mass of chinstrap males relative to adelie
# excess mass of gentoo males relative to adelie
```


- Using only the coefficients, what is the expected mass of male and female Gentoos with each model.

```{r}
# no interaction female
coef(mod1)["(Intercept)"] + coef(mod1)["speciesGentoo"]
# no interaction male
coef(mod1)["(Intercept)"] + coef(mod1)["sexmale"] + coef(mod1)["speciesGentoo"]


# interaction female
coef(mod2)["(Intercept)"] + coef(mod2)["speciesGentoo"]
# interaction male
coef(mod2)["(Intercept)"] + coef(mod2)["sexmale"] + coef(mod2)["speciesGentoo"] + coef(mod2)["sexmale:speciesGentoo"]

# check
nd <- tibble(sex = c("female", "male"), species = "Gentoo")
augment(mod1, newdata = nd)

augment(mod2, newdata = nd)

```

- Examine the models' diagnostic plots

```{r}
#| label: fig-diagnostics
#| fig-height: 10
performance::check_model(mod1)
performance::check_model(mod2)
# look fine
```


- Test if the interaction improves the model.
```{r}
anova(mod1, mod2) # p < 0.05
```


# Task 2 ancova

- With the Gentoo data, fit models to predict bill length from body mass and sex with and without an interaction.

```{r}
# filter Gentoo data
gentoo <- penguins |> filter(species == "Gentoo")


# model without interaction
mod3 <- lm(bill_length_mm ~ body_mass_g + sex, data = gentoo)
#model with interaction
mod4 <- lm(bill_length_mm ~ body_mass_g * sex, data = gentoo)
```

- Interpret the model coefficients.
```{r}
tidy(mod3)
# bill length for females when body mass = 0
# slope - 1 g extra body mass gives 0.0027 mm extra bill length
# offset for males

tidy(mod4)
# bill length for females when body mass = 0
# slope for females - 1 g extra body mass gives 0.0019 mm extra bill length
# offset for males at body mass = 0
# difference in slope for males 0.0014 mm/g steeper

```



- Using only the coefficients, what is the expected bill length of male and female Gentoos with body mass of 5000 g with each model.

```{r}
# no interaction
#female
coef(mod3)["(Intercept)"] + coef(mod3)["body_mass_g"] * 5000
#male
coef(mod3)["(Intercept)"] + coef(mod3)["body_mass_g"] * 5000 + coef(mod3)["sexmale"]

# interaction
#female
coef(mod4)["(Intercept)"] + coef(mod4)["body_mass_g"] * 5000
#male
coef(mod4)["(Intercept)"] + coef(mod4)["body_mass_g"] * 5000 + coef(mod4)["sexmale"] + coef(mod4)["body_mass_g:sexmale"] * 5000


#check
nd <- tibble(sex = c("female", "male"), body_mass_g = 5000)
augment(mod3, newdata = nd)
augment(mod4, newdata = nd)
```



- Examine the models' diagnostic plots

```{r}
performance::check_model(mod3)
performance::check_model(mod4)
```

- Plot the raw data and the models, including confidence intervals.

```{r}
augment(mod3, interval = "confidence") |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, colour = sex)) +
  geom_point() +
  geom_ribbon(aes(ymax = .upper, ymin = .lower, group = sex), colour = NA, alpha = 0.3) +
  geom_line(aes(y = .fitted))

augment(mod4, interval = "confidence") |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, colour = sex)) +
  geom_point() +
  geom_ribbon(aes(ymax = .upper, ymin = .lower, group = sex), colour = NA, alpha = 0.3) +
  geom_line(aes(y = .fitted))
```



- Test if the interaction improves the model.

```{r}
anova(mod3, mod4) # p > 0.05 - not significantly better
```


# Task 3

- Use the MuMIn package to find the "best" model for predicting body mass of Adelie penguins.
- Is there a clear best model, or several good models?

```{r}
adelie <- penguins |> filter(species == "Adelie") |> 
  drop_na() |> #remove ALL NA values
  select(-species) # now only one species, so cannot use as a predictor
mod5 <- lm(body_mass_g ~ ., data = adelie, na.action = na.fail) # have to use na.fail or dredge fails
library(MuMIn)
dredge(mod5) #top few models have similar AICc
```


