---
title: "Practical 5"
author: "Richard Telford"
date: "9/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#activate packages - best done in setup chunk to avoid startup messages being printed
library(tidyverse)
# load data  - or use library(palmerpenguins)
data(penguins, package = "palmerpenguins")

# set default theme
theme_set(theme_bw())
```

Do all the exercises in R markdown.

## Describing penguins

With the penguin data

-  calculate the minimum, maximum and range of bill lengths for Gentoo penguins

```{r gentoo-min}
gentoo <- penguins |> 
  filter(species == "Gentoo")

min(gentoo$bill_length_mm, na.rm= TRUE)
# OR
gentoo |> 
  drop_na(bill_length_mm) |> 
  summarise(
    min_bill_len = min(bill_length_mm),
    max_bill_len = max(bill_length_mm),
    range = max_bill_len - min_bill_len
  )
```

- calculate the mean and median body mass for each species

```{r}
penguins |> 
  drop_na(body_mass_g) |>
  group_by(species) |> 
  summarise(
    mean_mass = mean(body_mass_g),
    median_mass = median(body_mass_g)
  )

```

- calculate the variance and the standard deviation of body mass for each species.
```{r}
penguins |> 
  drop_na(body_mass_g) |>
  group_by(species) |> 
  summarise(
    var_mass = var(body_mass_g),
    sd_mass = sd(body_mass_g)
  )


```


- Calculate the standard error of the mean for the body mass of each species.

```{r se-mass}
penguins |> 
  drop_na(body_mass_g) |>
  group_by(species) |> 
  summarise(
    sd_mass = sd(body_mass_g),
    n = n(),
    se_mass = sd_mass/sqrt(n)
  )


```

- Calculate the 95% confidence interval for the mean body mass.

```{r ci-mass}
penguins |> 
  drop_na(body_mass_g) |>
  group_by(species) |> 
  summarise(
    mean_mass = mean(body_mass_g),
    sd_mass = sd(body_mass_g),
    se_mass = sd_mass/sqrt(n()),
    upper_ci = mean_mass + 1.96 * se_mass,
    lower_ci = mean_mass - 1.96 * se_mass
  )


```

## We want to test the hypothesis that bill length is sex dependent in Adelie penguins.

- What is the null hypothesis
- Make a plot of the relevant data
- Choose a suitable statistical test and run it. 
- Interpret the output.

```{r adelie}
adelie <- penguins |> 
  filter(species == "Adelie") |> 
  drop_na(sex, bill_length_mm)
```

```{r adelie-plot}

adelie |> 
  ggplot(aes(x = sex, y = bill_length_mm, fill = sex)) +
  geom_violin(draw_quantiles = 0.5, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(x = "Sex", y = "Bill length mm")
  
  
```

```{r adelie-t-test}
t.test(bill_length_mm ~ sex, data = adelie)
```



## Power test

We want to test whether fertiliser A is better than fertiliser B gives a higher yield of tomatoes.

- A typical tomato plant yields 1 kg of tomatoes (sd = 0.2 kg)
- A change of yield of 0.1 kg would be interesting

- Design an experiment to test this hypotheis
- How many replicates do we need to have an 80% probability of detecting an effect of 0.1kg at p = 0.05 
- If we want stronger evidence (p = 0.01), how many replicates do we need.

- hint: `pwr` package

```{r}
library(pwr)
#need a t-test power analysis
tomato_power <- pwr.t.test(
  d = 0.5, # difference in mean (0.1)/ sd (0.2)
  sig.level = 0.05, #typical value
  power = 0.8 # typical value
)

tomato_power
plot(tomato_power)
```


Also explore this graphically with simulations using `biostats.tutorials::power_lm_app()` **You will need to install the biostats.tutorial package again** 

For the exercise below, decide from the description of the hypothesis and data what type of analysis would be appropriate. 

## Exercise 3.

You have a dataset that is about size of bilberry leaves depending on light level in the area they have 
been growing. Each observation (leaf) is assumed to be independent of the others, i.e. there are no 
clustering of the data. The response variable (Leaf.area) is a continuous variable. The predictor variable
(Light.level) is a categorical variable with the three levels; low, intermediate and high.
You want to test the following hypothesis: H0: The leaf area of bilberries (cm) does not depend on light 
level in the area where the plants have been growing.
What family of test would you choose for this kind of data and what R function should you use? 

- independent obs
- response variable continuous
- predictor variable categorical
- linear model (anova) lm()

## Exercise 4.

You suspect that the number of eggs in the brood chamber of _Daphnia magna_ varies
depending on the protein availability of the feed when the mother is producing her eggs. You
have done an experiment where you have used two different feed qualities with respect to
amount of proteins available. You did the experiment to test the following hypothesis:
H0: The number of eggs in the third clutch of _D. magna_ mothers do not depend on protein availability 
of the feed.
You count the number of eggs from 30 females in each of the two feed groups. The observations are 
assumed to be independent of each other since the _D. magna_ mothers were kept singly in jars.

What family of test would you choose for this kind of data and what R function should you use?

- observations independent
- response counts
- predictor categorical
- gemeralised linear model (poisson family) glm()

## Exercise 5.

When working in field you discovered that the zygoptera larvae in five lakes in Bergen had a new type 
of ectoparasite never described before. As a first step in your investigations about this parasite, you 
want to compare the probability of being infested depending on lake. You did the study in a simple way
by collecting 30 individual zygoptera larvae in each lake and checked whether they were infested 
(response variable given the value 1) or not (response variable given the value 0).
You want to test the following hypothesis: H0: The probability of infestation by the ectoparasite do not 
differ between lakes.

What family of test would you choose for this kind of data and what R function should you use?

- observations independent
- response binary
- predictor categorical
- generalised linear model (binomial family) glm()
- you would never get funding for this work - it is a very uninteresting hypothesis
